Bootstrapping Domain Specific Languages for Neurally-Guided Bayesian Program Learning

### 2018-02-09 paper.tex
initial DSL primitives and types
restyled induced primitives

To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar,
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
a DSL in \system may be context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Supplement we derive a tractable MAP estimator for $\theta$.
%% \begin{algorithm}[tb]
%%    \caption{DSL Induction Algorithm}
%%    \label{grammarInductionAlgorithm}
%%    \begin{algorithmic}
%%      \STATE {\bfseries Input:} Set of frontiers $\{\mathcal{F}_x\}$
%%      \STATE \textbf{Hyperparameters:} Pseudocounts $\alpha$, regularization parameter $\lambda$
%%      \STATE \textbf{Output:} DSL $\mathcal{D}$, weight vector $\theta$
%%      \STATE Define $L(\mathcal{D},\theta) =  \prod_x \sum_{z\in \mathcal{F}_x} \probability[z|\mathcal{D},\theta]$
%%      \STATE Define $\theta^*(\mathcal{D}) = \argmax_\theta \text{Dir}(\theta|\alpha) L(\mathcal{D},\theta)$
%%      \STATE Define $\text{score}(\mathcal{D}) = \log \probability[\mathcal{D}] + L(\mathcal{D},\theta^*) - ||\theta||_0$
%%      \STATE $\mathcal{D}\gets$ every primitive in $\{\mathcal{F}_x\}$
%%      \WHILE {true}
%%      \STATE N $\gets \{\mathcal{D}\cup \{s\} | x\in X, z\in \mathcal{F}_x, s\text{ subexpression of }z\}$
%%      \STATE $\mathcal{D}'\gets \argmax_{\mathcal{D}'\in N}\text{score}(\mathcal{D}') $
%%      \STATE \textbf{if }$\text{score}(\mathcal{D}') < \text{score}(\mathcal{D})$\textbf{ return }$\mathcal{D},\theta^*(\mathcal{D})$
%%      \STATE $\mathcal{D}\gets\mathcal{D}'$
%%      \ENDWHILE
%%    \end{algorithmic}
%% \end{algorithm}

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
    $f_0 = $\code{(+ }$\mathcal{R}$\code{)}\\
    $f_1($\code{x}$) =$ \code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)} \\
      $f_2($\code{x}$) =$ \code{(+ }$\mathcal{R}$\code{ ($f_1$ x))} \\
      $f_3($\code{x}$) = $\code{(* x (}$f_2$\code{ x))}\\
  $f_4($\code{x}$) = $\code{(}$f_0$\code{ (}$f_3$\code{ x)))}\\
  $f_5($\code{x}$) =$ \code{(* x (}$f_4$\code{ x))}\\
  $f_6($\code{x}$) = $\code{(}$f_0$\code{ (}$f_4$\code{ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{+}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) = $\code{(+ (chr->str c))}\\
      $f_1($\code{i,c,s}$) = $      \code{(+ (nth i (split c s))}\\\hspace{2.55cm}\code{ (chr->str c))}\\
      $f_2($\code{s,c}$) = $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
      \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c,s}$) = $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
      \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}

@@

  \begin{minipage}{7.5cm}
 $f($\code{s}$) = $   \code{(}$f_3$\code{ (split '>' (upper s)) '<')}
    \end{minipage}
  \caption{Example string edit task (top) and program  \system writes (bottom). $f_3$ is a new DSL subroutine discovered by \systemEnding, defined in Tbl.~\ref{textPrimitives}}\label{exampleTextProblem}
  \end{figure}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{l}
    \toprule
    $(f_0$ \code{i $\ell$}$) = $\code{(singleton (index i $\ell$))}\\
    \hspace{0.5cm}($f_0$: \emph{Put the $i$-th index into a list})\\
    $(f_1$ \code{i $\ell$}$) = $\code{(++ $\ell$ ($f_0$ i $\ell$))}\\
    \hspace{0.5cm}($f_1$: \emph{Append the $i$-th index})\\
    $(f_2$ \code{n $\ell$}$) = $\code{(any (lambda (x) (= n x)) $\ell$)}\\
    \hspace{0.5cm}($f_1$: \emph{Whether $n$ appears in the list})\\
    $(f_3$ \code{i $\ell$}$) = $\code{(index (negate i) (sort $\ell$))}\\
    \hspace{0.5cm}($f_3$: \emph{Get the $i$-th largest number})\\
    $(f_4$ \code{n $\ell$}$) = $\code{(mapi (lambda (i x) (mod x n)) $\ell$)}\\
    $(f_5$ \code{k $\ell$}$) = $\code{(mapi (lambda (i x) (+ x k)) $\ell$)}\\
    $(f_6$ \code{k n $\ell$}$) = $\code{($f_4$ n ($f_5$ k $\ell$)}\\
    \hspace{0.5cm}($f_6$: \emph{Caesar shift of $k$ in integers modulo $n$})\\
  \bottomrule
  \end{tabular}
  \caption{Some induced primitives for list functions.}\label{listinduced}
\end{table}

>>>>

To appropriately score each proposed $\mathcal{D}$ we must reestimate the weight vector $\theta$.
Although this problem may seem 
very similar to estimating the parameters of a probabilistic context free grammar,
for which we have effective approaches like the Inside/Outside algorithm~\cite{international2000derivation},
a DSL in \system may be context-sensitive due to the presence of variables
in the programs and also due to the polymorphic typing system.
In the Supplement we derive a tractable MAP estimator for $\theta$.

@@

What DSL does \system learn?
The learned DSL contains templates for polynomials of different orders,
which lets the algorithm quickly hone in on the kinds of functions that are most appropriate to this domain (Tbl.~\ref{regressionDSL}).
Examining the programs themselves,
one finds that the algorithm discovers representations for each of the polynomials that minimizes the number of continuous degrees of freedom:
for example, it represents the linear function $-x+2$ with the program 
\code{(* (* (+ x }$\mathcal{R}$\code{) }$\mathcal{R}$\code{) x)}, which has two continuous degrees of freedom -- while representing quartic functions using the invented DSL primitive $f_6$ in Tbl.~\ref{regressionDSL}.
This falls out of our Bayesian framing -- both the bias towards shorter programs (more likely under $\mathcal{D},\theta$) and the likelihood model's BIC penalty.
\begin{table}\centering
\begin{tabular}{l}
  \toprule
  %% Primitives& $+,\times:\mathbb{R}\to \mathbb{R}\to \mathbb{R}$\\
  %% & $\mathcal{R}:\mathbb{R}$ (real valued parameter)\\\midrule
  %% Observation $x$& $N$ input/output examples: $\left\{(i_n,o_n) \right\}_{n\leq N}$\\
  %% \midrule
  %% Likelihood $\probability[x|p]$& $\propto\exp\left( - D\log N \right)\prod_{n\leq N}\mathcal{N}(p(i_n)|o_n)$
  %% \\\midrule
  $f_0\,=\,$\code{(+ $\mathcal{R}$)}\\
  $f_1($\code{x}$)\,=\,$\code{(* (* (+ x $\mathcal{R}$) $\mathcal{R}$) x)} \\
  $f_2($\code{x}$)\,=\,$\code{(+ $\mathcal{R}$ ($f_1$ x))} \\
  $f_3($\code{x}$)\,=\,$\code{(* x ($f_2$ x))}\\
  $f_4($\code{x}$)\,=\,$\code{($f_0$ ($f_3$ x))}\\
  $f_5($\code{x}$)\,=\,$\code{(* x ($f_4$ x))}\\
  $f_6($\code{x}$)\,=\,$\code{($f_0$ ($f_4$ x))} (\emph{4th order polynomial})
  \\\bottomrule
\end{tabular}
\caption{Some induced DSL primitives for symbolic regression. System starts out with addition, multiplication, and real numbers, and learns to build 4th order polynomials.}\label{regressionDSL}
\end{table}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{c}
    String editing primitives\\
    \begin{tabular}{l}
    \toprule
    \code{0}, \code{+1},\code{-1},     \code{""},    \code{","},     \code{" "},    \code{"<"}, \code{">"}  \\
    \code{split}, \code{join}, \code{map}, 
    \code{+}, \code{chr->str}, \code{trim}, \code{nth}\\
    \code{length}, \code{slice},    \code{lower}, \code{upper}, \code{capitalize}
  \\\bottomrule
    \end{tabular}\\
\\    Subset of learned DSL\\
    \begin{tabular}{l}
      \toprule
      $f_0($\code{c}$) \,=\, $\code{(+ (chr->str c))}\\
      $f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ (nth i (split c s))}\\
        \phantom{$f_1($\code{i$,$c$,$s}$) \,=\, $\code{(+ }}\code{(chr->str c))}\\
      $f_2($\code{s$,$c}$) \,=\, $\code{((}$f_0$\code{ c) (}$f_1$\code{ (1+ 0) c s))} \\
        \hspace{0.5cm}($f_2$: \emph{Extract substring delimited by character})\\
      $f_3($\code{c$,$s}$) \,=\, $\code{(nth (1+ 0) (split c (nth 0 s)))}\\
        \hspace{0.5cm}($f_3$: \emph{2nd substring delimited by character})
    %% $f_0($\code{i,c,s}$) = $    \code{(+ (index i (split c s)) (chr->str c))}\\
    %% $f_1 = $\code{(join "")}
  \\\bottomrule
    \end{tabular}
  \end{tabular}
  \caption{Top: provided string editing primitives. Bottom: some induced DSL primitives.}\label{textPrimitives}
  \end{table}

@@

  \begin{minipage}{7.5cm}
    $f($\code{s}$) \,=\, $\code{(}$f_3$\code{ (split '>' (upper s)) '<')}
  \end{minipage}
  \caption{A string edit task and the program \system writes for it. $f_3$ is a discovered DSL subroutine, defined in Tbl.~\ref{textPrimitives}}\label{exampleTextProblem}.
  \end{figure}

@@

\begin{table}[t]
  \centering
  \begin{tabular}{l}
    \toprule
    $f_0($\code{i$,\ell$}$) \,=\, $\code{(singleton (index i $\ell$))}\\
    \hspace{0.5cm}($f_0$: \emph{Put the $i$-th index into a list})\\
    $f_1($\code{i$,\ell$}$) \,=\, $\code{(++ $\ell$ ($f_0$ i $\ell$))}\\
    \hspace{0.5cm}($f_1$: \emph{Append the $i$-th index})\\
    $f_2($\code{n$,\ell$}$) \,=\, $\code{(any (lambda (x) (= n x)) $\ell$)}\\
    \hspace{0.5cm}($f_1$: \emph{Whether $n$ appears in the list})\\
    $f_3($\code{i$,\ell$}$) \,=\, $\code{(index (negate i) (sort $\ell$))}\\
    \hspace{0.5cm}($f_3$: \emph{Get the $i$-th largest number})\\
    $f_4($\code{n$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (mod x n)) $\ell$)}\\
    $f_5($\code{k$,\ell$}$) \,=\, $\code{(mapi (lambda (i x) (+ x k)) $\ell$)}\\
    $f_6($\code{k$,$n$,\ell$}$) \,=\, $\code{($f_4$ n ($f_5$ k $\ell$)}\\
    \hspace{0.5cm}($f_6$: \emph{Caesar shift of $k$ in integers modulo $n$})\\
  \bottomrule
  \end{tabular}
  \caption{Some induced primitives for list functions.}\label{listinduced}
\end{table}
